---
title: "Example R Markdown with an analysis of Iris data"
author: "Wiesław Babik"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
```

# Load and summarise

We'll use built-in `iris` dataset containing various flower measurements in three *Iris* species.

First, we have a look at the structure of the data

```{r}
str(iris)
```

Second, we print part of the data as a simple table 

```{r}
#we use kable() from knitr package
kable(head(iris))
```

Third, we generate basic summary of the dataset and print it as a simple table

```{r}
kable(summary(iris))
```

# Visualise and interpret

Finally, we plot the relationship between sepal length and sepal width and interpret results with the help of correlations coefficients calculated as inline R code.

## Plot

```{r}
p <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, colour = Species)) +
  geom_point() + geom_smooth(method = 'lm') +
  theme_light()
p

```

## Interpretation

We first calculate Pearson correlation coefficient for each species and then write a short interpretative text 

```{r}
r <- iris %>% group_by(Species) %>% 
  summarise(r = round(cor(Sepal.Length, Sepal.Width, method = 'pearson'), 2)) %>% 
  deframe()

```

It looks like correlation for *setosa* (`r r["setosa"]`) is higher than for the two other species (`r paste(r[[2]], r[[3]], sep =", ")`), but we have to do proper statistical test for significance of the difference.


